Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-08 summary.tex
starting to edit summary (not done yet)

\section{Summary and discussion}\label{summary}
In this paper we have studied the axisymmetric vertical shear 
instability (VSI) in astrophysical disks that arise from the 
height dependence of the equilibrium angular velocity profile,  
$\p_z\Omega\neq 0$. Vertical shear is generally
expected in astrophysical disks due to baroclinicity ($\nabla
P\times\nabla \rho \neq 0$). This may be due
to a radial variation in the disk temperature, or more generally in
the disk entropy. However, astrophysical disks are also typically
sub-adiabatically stratified for stability against vertical
convection ($N_z^2\geq 0$). In this case, the VSI requires a short
thermal or cooling timescale $t_c$ in order to operate.

>>>>

\section{Summary and discussion}\label{summary}
In this paper we study the vertical shear instability (VSI) with a focus on 
the radiative cooling requirements.  In turn we address the viability of the VSI 
as an angular momentum transport mechanism in protoplanetary disks (PPDs).

Our linear axisymmetric analysis of the VSI 
considers (uniquely to our knowledge) finite cooling times in a vertically global model.  
Our main analytical finding, which we confirm numerically, is the critical cooling 
timescale above which VSI growth is suppressed, Eqs.\ \ref{prelim_bcrit} and {iso_vsi_cond}.
Short cooling times are needed to weaken the stabilizing effects of vertical buoyancy, thereby allowing 
the free energy in vertical shear to drive instability.

Our main focus is irradiated, vertically isothermal disks which have strong vertical buoyancy.
The critical cooling time is thus short, shorter than the orbital time by a factor of the disk
aspect ratio.  This finding is consistent with, and helps explain, the results of 3D numerical simulations \citepalias{N13}.

In applying our results to PPDs, we pay particular attention to the transition from 
optically thick radiative diffusion to optically thin Newtonian cooling.  The largest obstacle to VSI occurs in 
high density inner disk, $\lesssim 1$ -- 5 AU, where radiative diffusion times are slow.  Shorter wavelength disturbances
 cool faster, though only if they remain optically thick.  In the inner disk cooling can only be fast enough
 if wavelengths are too short to drive significant transport.  The best hope for the VSI in inner disk is a low surface density.

...

In this paper we have studied the axisymmetric vertical shear 
instability (VSI) in astrophysical disks that arise from the 
height dependence of the equilibrium angular velocity profile,  
$\p_z\Omega\neq 0$. Vertical shear is generally
expected in astrophysical disks due to baroclinicity ($\nabla
P\times\nabla \rho \neq 0$). This may be due
to a radial variation in the disk temperature, or more generally in
the disk entropy. However, astrophysical disks are also typically
sub-adiabatically stratified for stability against vertical
convection ($N_z^2\geq 0$). In this case, the VSI requires a short
thermal or cooling timescale $t_c$ in order to operate.
